---
title: "Data visualization assignment"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, fig.width = 6, fig.height = 4)
```

In this exercise, you will practice basic plotting with **`ggplot2`** on a
dataset of median gene expression levels across samples from various human
tissues from the GTEx Project. The dataset (`dt`) has one row per gene, with
columns:

-   `Name`: the Ensembl gene ID
-   `Description`: the gene symbol
-   One column per tissue containing the median expression levels (in TPM units)

Insert your code into the empty code blocks. Insert answers to any comprehension
questions as comments within the code blocks. You do not need to submit the
figures themselves, only the code used to produce them. Remember to label your
axes!

**Using HopGPT.** As in the last assignment: you are welcome to use it, and
wherever you do, add a couple of comment lines in that same code block noting
what you asked it and how you checked the answer.

------------------------------------------------------------------------

### Load data

Start by loading the data into R, as before.

```{r load-data}
library(tidyverse)

dt <- read_delim("GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

Recreate the `is_mitochondrial` column from the last assignment; you will need it
in Q4 and Q5.

```{r add-mito}
dt <- dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-"))

dt %>% count(is_mitochondrial)
```

------------------------------------------------------------------------

### Q1. Histogram

Make a histogram of gene expression in the liver. Use the argument
`breaks = seq(0, 100, 2)` to bin in units of 2 TPM, ranging from 0 to 100. Note
that this cuts off the data at 100, which might be misleading depending on what
we are trying to depict! How many genes are you leaving out?

What do you notice about the shape of the distribution? What does it tell you
about the relative expression of genes in the liver transcriptome? Do you think
this is specific to the liver? Check another tissue, as well.

```{r q1}

# answer here
```

------------------------------------------------------------------------

### Q2. Density plot

Going back to the liver, now represent the same data with a density plot rather
than a histogram. Do not restrict the X axis. Within the same code block, create
a second plot where you scale the X axis to a log10 scale. Add a pseudo-count of
1 to the expression values, as the log of 0 is negative infinity.

What does the log-scaled version show you that the first one does not? And where
on the log-scaled plot do the genes with 0 TPM end up?

```{r q2}

# answer here
```

------------------------------------------------------------------------

### Q3. Scatter plot

Make a scatter plot comparing the expression of genes in the `Liver` to the same
genes in the `Heart - Left Ventricle`. Add a pseudocount to each and log10 scale
both the X and Y axes. Adjust the point size to 1 and the alpha (opacity) to 0.2.
Add a dashed red line along the diagonal (y = x).

What does a point far above the diagonal mean, biologically?

```{r q3}

# answer here
```

------------------------------------------------------------------------

### Q4. Colored points on scatter plot

Color the previous scatter plot points based on whether the genes are
mitochondrial vs nuclear encoded, using the `is_mitochondrial` column you made
above.

There are only a few dozen mitochondrial genes among tens of thousands of
nuclear ones, so the colored points end up buried. Try plotting them as a second
`geom_point()` layer on top of the others, using the `data` argument to give that
layer just the mitochondrial genes:
`geom_point(data = filter(dt, is_mitochondrial), ...)`.

Where do the mitochondrial genes sit relative to the diagonal, and does that fit
what you found in the last assignment?

```{r q4}

# answer here
```

------------------------------------------------------------------------

### Q5. Facets

Repeat question 1 (histogram), but faceting such that you create separate panels
(arranged horizontally) for several tissues of interest:

```{r focal-tissues}
focal_tissues <- c("Liver", "Testis", "Whole Blood", "Heart - Left Ventricle", "Brain - Cortex")
```

This requires first converting the data into "long" format with `pivot_longer()`:

```{r pivot-long}
dt_long <- dt %>%
  select("Description", "is_mitochondrial", all_of(focal_tissues)) %>%
  pivot_longer(cols = all_of(focal_tissues), names_to = "Tissue", values_to = "TPM")
```

After creating the 5-panel plot, also split into two rows that separate genes
that are versus are not encoded by the mitochondrial genome (10 panels). Use the
argument `scales = "free_y"` to deal with the fact that there are far fewer
mitochondrial-encoded than nuclear-encoded genes.

`scales = "free_y"` means the panels can no longer be compared by eye. Write the
one-sentence note you would put in a figure caption to keep a reader from being
misled.

```{r q5}

# your code for 5-panel plot

# your code for 10-panel plot

# caption note

```

------------------------------------------------------------------------

### Q6. Write a function

By now you have written nearly the same histogram several times, changing only
the tissue. Write a function that does it once.

Your function should take a vector of expression values and a tissue name, and
**return** a log10-scaled histogram of those values (with a pseudocount), with
the tissue name in the title and both axes labeled. Give the pseudocount a
default value of 1.

As in the live coding, have the function take a plain vector, which you pull out
with `$`, so `plot_tissue_histogram(dt$Liver, "Liver")` is how you would call it.

Test it on the liver and confirm you get the same figure you made in Q2 --- that
is the point of the test. Then use it to produce the plot for each of the five
`focal_tissues`, either with a `for` loop or by calling it five times.

```{r q6}

# your function here

# test it against a result you already trust

# all five tissues

```

------------------------------------------------------------------------

Optional: Have fun and explore these data! Create your own plots and interpret
them.

```{r optional}

# answer here
```
